---
title: "Initial Report - Diatoms"
output: pdf_document
authors: Sydney Barber and Gregory J. Matthews
date: "2026-04-20"
---

# Summary 
To clean the data, we removed the major groups that had unknown or unidentified groups. 
Then it transforms the data from wide to long format, where all individual element names are moved into a single column and their corresponding values into another. Finally, it replaces any missing values with zeros and calculates the total counts, grouped by variables like depth, sample number, and duration.

The graphs below shows the counts for the different major groups through time. It is important to note that the scales on the Y-axis are different for the different groups. The use of free scales in the plot allows for the detection of relative population dynamics within less abundant groups, such as F1 or GSSC, which would otherwise be flattened by the higher magnitudes of dominant groups like Centric diatoms. This visualization serves as a diagnostic tool to identify synchronous shifts across different taxa, suggesting potential environmental thresholds or 'tipping points' in the depositional history. The groups that seem to be the most prevelant are Centric, NDO, and Pennate.

Initially it seems that there is a trend in F1, GSSC, Biogenic, and NDG. However, statisitcal analysis must be done to prove this.

```{r setup, include=FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
taxon <- read.csv("./data/taxonomy_species_table.csv")
final <- read.csv("./data/Final_diatom_phytolith_by_transect.csv")
```

```{r, echo = FALSE, message = FALSE}
final <- final %>% 
  dplyr::select(-any_of(c(
    paste0("Diatom_Unidentified_", LETTERS[1:5]),
    "Diatom_Unknown_Diatom",
    "Diatom_Unknown_Diatom.fragment",
    "Others_Granules",
    paste0("Others_Undetermined_", LETTERS[7:12])
  )))

finallong <- final %>% pivot_longer(cols = Diatom_Symmetric_Biraphid_spp.:Others_Charred, names_to = "Elements", values_to = "Count") 
finallong$Count[is.na(finallong$Count)] <- 0
test <- finallong %>% group_by(D_level,sample_num,  Depth..m.,Duration,Elements,) %>% summarize(Count = sum(Count))

```

```{r echo = FALSE, message = FALSE}
summary_data <- finallong %>%
  group_by(Mean.time.of.deposition, D_level) %>%
  summarise(count = sum(Count), .groups = "drop")

test2 <- finallong %>% group_by(D_level,sample_num, Mean.time.of.deposition, Depth..m.,Duration,Elements,) %>% summarize(Count = sum(Count))

unique <- unique(test2$Elements)

taxon <- taxon %>% filter(species %in% unique)

combined_data <- test2 %>%
  left_join(taxon, by = c("Elements" = "species"))


d_order <- paste0("D", 1:21)

combined_data <- combined_data %>%
  mutate(D_level = factor(D_level, levels = d_order))

plot_data <- combined_data %>%
  filter(major_group != "Unidentified") %>%
  group_by(major_group, Mean.time.of.deposition) %>%
  summarise(total_val = sum(Count, na.rm = T), .groups = "drop") 


ggplot(plot_data %>% filter(!is.na(Mean.time.of.deposition)), aes(x = Mean.time.of.deposition, y = total_val,  group = major_group)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(
    title = "Total Counts by Major Group across time points",
    x = "Mean time of deposition",
    y = "Sum of Counts",
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~major_group, scales="free") +
  geom_smooth()
```


<!-- Greg's Notes/What he Wants To Do -->


<!-- # #KENDALLS TOW (no assumptions of linearity)-concordance -->
<!-- # #spearman  -->
<!-- #  -->
<!-- # #Stratify by group -->
<!-- #  -->
<!-- # #Test for ANY dependence.   -->
<!-- # #Citation: https://projecteuclid.org/journals/annals-of-statistics/volume-35/issue-6/Measuring-and-testing-dependence-by-correlation-of-distances/10.1214/009053607000000505.full -->
<!-- # library(energy) -->
<!-- # dcor.test(plot_data$Mean.time.of.deposition, plot_data$total_val) -->
<!-- #  -->
<!-- # #Test for monotonic relationship -->
<!-- # cor.test(plot_data$Mean.time.of.deposition, plot_data$total_val, method = "kendall") -->


# Statisitcal Analysis
```{r, echo=FALSE, message=FALSE}
library(energy)
set.seed(413)
run_dependency_tests <- function(df)  {
  
  #Test Greg Found which tests for any dependence
  dcor_res <- energy::dcor.test(df$Mean.time.of.deposition, df$total_val, R = 1000) 
  #R is permutations, how many should we do
  
  #Kendall's Tau 
  ken_res <- cor.test(df$Mean.time.of.deposition, df$total_val, method = "kendall", exact = F)
  
  #Returning the data in a clean matter
  tibble(
    dcor_stat = dcor_res$statistic,
    dcor_p    = dcor_res$p.value,
    kendall_tau = ken_res$estimate,
    kendall_p   = ken_res$p.value
  )
}

test_results <- plot_data %>%
  filter(!is.na(Mean.time.of.deposition)) %>%
  group_by(major_group) %>%
  nest() %>%
  mutate(tests = map(data, ~run_dependency_tests(.x))) %>%
  unnest(tests) %>%
  ungroup() %>%            
  dplyr::select(-data)




test_results <-test_results %>% 
  mutate(kendall_p_adj = p.adjust(kendall_p, method="BH"),
         dcor_p_adj = p.adjust(dcor_p, method ="BH" )
         ) %>% select (major_group, dcor_stat, dcor_p, dcor_p_adj, kendall_tau, kendall_p, kendall_p_adj)




library(kableExtra)

kable(test_results, caption = "Correlation Test Results") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE)


```

We ran Kendall's Tau and a Distance Correlation Test (DCT) on each of the major groups. We then adjusted for multiple hypothesis testing using the Benjamini-Hochberg (BH) False Discovery Rate Correction. The DCT tests the null hypothesis of independence between x and y, which in this case is time and counts of each of the major groups at that specific time point versus the alternative of not independence: dependence. Based on the results of this test, the groups Biogenic and FI rejected the null hypothesis at the 0.05 level meaning that there is detected significally significant evidence of dependence. Then, for the Kendall's Tau Test we tested the null hypothesis that Kendall's Tau is 0, versus the alternative that it is not 0. Based on the results of this test, the group Biogenic rejected the null hypothesis at the 0.05 level meaning that there is statistically significant evidence of a discordant relationship between time and the count of the group. It is interesting to note that Biogenic was rejected in both tests which indicates that there is strong evidence of a relationship.
